ScPy: A SuperCollider Extension for Performing Numerical Computation via Embedded Python

### 2016-08-29 doc/paper.tex
shuffling, editing, conclusion

ScPy enables users to overcome these limitations. By embedding the Python programming language
within SuperCollider, it is possible to do numerical processing operations which would be too slow
to perform in real time with SuperCollider alone. Through access to the NumPy library, previously
difficult or impossible phase vocoder operations become trivial. Although our usage of this library
for testing purposes has mostly focused on FFT based operations, support exists for passing
arbitrary data buffers to Python, which can be used to process numerical data for many other
purposes as well.

ScPy is developed and maintained as an open-source project under the GPLv3 license. The latest
code, documentation, and issues are available on
GitHub\footnote{\url{https://github.com/nwoeanhinnogaehr/ScPy}}. Contributions to the project are
welcome.

@@

ScPy is unlike many NIMEs because it does not have the properties of any traditional instrument.
However, it can still be considered a NIME because it ultimately is a tool for musical expression,
just with an unconventional method of interaction. A plot of ScPy in the 7-dimensional space
developed by Birnbaum et al.\ can be seen in~\autoref{fig:nime}.

\begin{figure}[ht]
    \caption{NIME dimension space analysis for ScPy.}\label{fig:nime}
    \begin{center}
        \begin{tikzpicture}[rotate=-116,xscale=-1]
            \tkzKiviatDiagram[label distance=.5cm,
                    radial=7,
                    gap=0.75,
                    lattice=4]{Required Expertise,
                    Musical Control,
                    Degrees of Freedom (Inputs),
                    Feedback Modalities (Outputs),
                    Inter-actors,
                    Distribution in Space,
                    Role of Sound}
            \tkzKiviatLine[thick,color=blue,mark=none,
                           fill=blue,opacity=.5](4,4,1.5,1.5,0.5,0.5,3)
        \end{tikzpicture}
    \end{center}
\end{figure}

@@

The phase vocoder currently implemented in ScPy uses the standard algorithm. Future experiments
could implement many of the proposed modifications to the phase vocoder to improve sound quality.
\\\par
Bradford et al.\ have published a number of works detailing the sliding
DFT~\cite{bradford2005sliding} and associated sliding phase vocoder~\cite{bradford2007sliding}.
These new algorithms are efficient implementations for the STFT with a hop size of a single sample.
Although such a small hop size would be impossible to process in real time using current hardware
and the STFT, the sliding DFT makes it possible using consumer GPU
hardware~\cite{bradford2011real}. The sliding phase vocoder additionally has a number of advantges
over traditional implementations. Quality is greatly improved (at a high computational cost). Each
analysis bin can hold any frequency, essentially making the spectrum more of a generic oscillator
bank than a traditional spectrum. This simplifies effects such as pitch shifting to a simple vector
multiplication in the frequency domain. Additionally, since the time resolution is so fine, effects
that require time modulation such as FM or AM can now be performed in the frequency domain.

Using Theano, the sliding DFT and sliding phase vocoder could be implemented easily in ScPy. It
remains unknown if it will be possible to run them in real time, but even if not this future
extension could add the possibility to experiment with many novel effects.
\\\par
Klingbeil~\cite{klingbeil2005software} describes the techniques used in a high quality graphical
spectral editing and resynthesis program called SPEAR\@. The FFT analysis is done taking into
account the fundamental frequency of the sound to analyze, so that the transform can be optimized
to accurately capture as many partials as possible. Spectral peaks are then selected and very quiet
peaks are discarded. A linear prediction method is used to follow the peaks over time. The data
structure for storage of spectral data involves division of time into a number of frames which keep
track of which partials are active during that time, so that the search for a specific partial can
be narrowed. User interface design choices are also covered in detail.

@@

As the Mellin transform can be implemented simply as an extension to the Fourier transform, it
should be fairly easy to add support for it in ScPy, and it will hopefully enable the use of
some different audio effects than can be achieved easily with the Fourier transform. However, the
paper mentions that a short-time Mellin transform has not been attempted, so some further work may
need to be done in that area first.
\\\par
Collins et al.~\cite{collins2003live} present an introduction to the field of live coding, where
code to generate sound is written live during a performance, often on a laptop computer. Although
the use graphical programming environments such as Reaktor, Pure Data, or MAX/MSP in live
performances is in some sense live coding, the paper focuses on the use of text based programming
environments. Many of the pros and cons of live coding are presented. Although live coding adds
some very interesting new possibilities to live music performance, it is not without it's
challenges. Case studies of two live coding environments are included: Slub and SuperCollider with
the just in time library.

ScPy can be used for live coding of audio effects, similarly to how the environments studied in
this paper can be used to live code synths and pattens. More details about how ScPy was designed to
facilitate this can be found in \autoref{sec:methodology}.

@@

Live coding presents an interesting contrast to contemporary music with many clear differences.
Nilson~\cite{nilson2007live} presents a review of the challenges of live coding
practice. While some argue that live coding must be performed live in front of an
audience, an alternate perspective suggests that time constraints are not necessary and that the
practice can refer to any improvisational music created by programming. The practice of live coding
is still quite young --- it is mentioned that professional violinists will typically have practiced
for 10000 hours, where as an email survey of live coders revealed most had dedicated about 100
hours. Of course, the prerequisite knowledge for live coding may be higher; many live coders have
extensive programming experience. Another thing mentioned is the high level of abstraction offered
by live coding creates a large startup time for producing meaningful output from a blank slate.
Additionally, many exercises are presented for improving one's live coding ability.

The ideas in Nilson's paper shed light on some of the challenges faced in this project.
One such challenge is the amount of writing required to produce an interesting result from a blank
slate. We intend to minimise this by providing many effect primitives and composition
mechanisms. However, it will be important to ensure the language is accessible to ensure it can be
quickly learned to a level of expertise where lack of technical knowledge is not an impediment.
\\\par
Domain-Specific Languages (DSLs) are small programming languages designed for use in a specific
problem domain. They often lack general purpose programming constructs and have a syntax which is
very concise for describing programs in their domain. Van Deursen et al.~\cite{van2000domain}
present a survey of available DSL literature, along with the pros and cons of using a DSL, example
DSLs, design methodology and implementation approaches. The biggest advantage of DSLs is that they
enable expression of solutions in a way that is common for the problem domain. However, they can be
costly to develop and difficult to design. Implementation techniques include the classical approach
of interpretation or compilation, embedded languages / domain-specific libraries, preprocessing, or
compiler extension. Additionally, the references of the paper each include a short summary.

@@

Use of ScPy is through the classes \texttt{Py} and \texttt{PyOnce}. \texttt{Py} is used for
processing real-time data, and \texttt{PyOnce} is used for doing initialization or other one time
processing. Both have one required argument, a block of Python code. A map of variable names may
optionally be provided to bind SuperCollider variables to Python variables. Additionally, there is
an optional \texttt{DoneAction} argument which specifies an action to occur in SuperCollider after
the Python code has finished executing.

@@

Since UGens may only be used as part of a synth, \texttt{PyOnce} is provided as a wrapper which
hides this detail from the user and allows the execution of Python code outside of a synth. It is
useful for doing initialization work. Since the current implementation executes Python code in a
single global namespace, a typical usage patten is for state to be defined in a \texttt{PyOnce}
block which is later accessed and modified in a \texttt{Py} block.

@@

Throughout~\autoref{sec:lit}, articles are discussed which contain processing techniques that could
be used in ScPy. Although implementations of many of these techniques are likely to become a part
of ScPy in the future, there is no obstacle to the reader implementing them as Python libraries
usable by ScPy or even as open-source contributions to the project.

\newpage\printbibliography{}

>>>>

ScPy enables users to overcome these limitations. By embedding the Python programming language
within SuperCollider, it is possible to do numerical processing operations which would be too slow
to perform in real-time with SuperCollider alone. Through access to the NumPy library, previously
difficult or impossible phase vocoder operations become trivial. Although our usage of this library
for testing purposes has mostly focused on FFT based operations, support exists for passing
arbitrary data buffers to Python, which can be used to process numerical data for many other
purposes as well.

ScPy is developed and maintained as an open-source project under the GPLv3 license. The latest
code, documentation, examples, and issues are available on
GitHub\footnote{\url{https://github.com/nwoeanhinnogaehr/ScPy}}. The code has been developed and
tested on the Linux operating system using SuperCollider version 3.6.6.

@@

ScPy is unlike many NIMEs because it does not have the properties of any traditional instrument.
However, it can still be considered a NIME because it ultimately is a tool for musical expression,
just with an unconventional method of interaction. A plot of ScPy in the 7-dimensional space
developed by Birnbaum et al.\ can be seen in~\autoref{fig:nime}.
\begin{figure}[ht]
    \caption{NIME dimension space analysis for ScPy.}\label{fig:nime}
    \begin{center}
        \begin{tikzpicture}[rotate=-116,xscale=-1]
            \tkzKiviatDiagram[label distance=.5cm,
                    radial=7,
                    gap=0.75,
                    lattice=4]{Required Expertise,
                    Musical Control,
                    Degrees of Freedom (Inputs),
                    Feedback Modalities (Outputs),
                    Inter-actors,
                    Distribution in Space,
                    Role of Sound}
            \tkzKiviatLine[thick,color=blue,mark=none,
                           fill=blue,opacity=.5](4,4,1.5,1.5,0.5,0.5,3)
        \end{tikzpicture}
    \end{center}
\end{figure}
\\\par
Live coding presents an interesting contrast to contemporary music with many clear differences.
Nilson~\cite{nilson2007live} presents a review of the challenges of live coding
practice. While some argue that live coding must be performed live in front of an
audience, an alternate perspective suggests that time constraints are not necessary and that the
practice can refer to any improvisational music created by programming. The practice of live coding
is still quite young --- it is mentioned that professional violinists will typically have practiced
for 10000 hours, where as an email survey of live coders revealed most had dedicated about 100
hours. Of course, the prerequisite knowledge for live coding may be higher; many live coders have
extensive programming experience. Also mentioned is the high level of abstraction offered
by live coding creates a large startup time for producing meaningful output from a blank slate.
Additionally, many exercises are presented for improving one's live coding ability.

The ideas in Nilson's paper shed light on some of the challenges faced in this project.
One such challenge is the amount of writing required to produce an interesting result from a blank
slate. We intend to minimise this by providing many effect primitives and composition
mechanisms. However, it will be important to ensure the language is accessible to ensure it can be
quickly learned to a level of expertise where lack of technical knowledge is not an impediment.

@@

The phase vocoder currently implemented in ScPy uses the standard algorithm. Future experiments
could implement many of the proposed modifications to the phase vocoder to improve sound quality.
\\\par
Bradford et al.\ have published a number of works detailing the sliding
DFT~\cite{bradford2005sliding} and associated sliding phase vocoder~\cite{bradford2007sliding}.
These new algorithms are efficient implementations for the STFT with a hop size of a single sample.
Although such a small hop size would be impossible to process in real-time using current hardware
and the STFT, the sliding DFT makes it possible using consumer GPU
hardware~\cite{bradford2011real}. The sliding phase vocoder additionally has a number of advantges
over traditional implementations. Quality is greatly improved (at a high computational cost). Each
analysis bin can hold any frequency, essentially making the spectrum more of a generic oscillator
bank than a traditional spectrum. This simplifies effects such as pitch shifting to a simple vector
multiplication in the frequency domain. Additionally, since the time resolution is so fine, effects
that require time modulation such as FM or AM can now be performed in the frequency domain.

Using Theano, the sliding DFT and sliding phase vocoder could be implemented easily in ScPy. It
remains unknown if it will be possible to run them in real-time, but even if not this future
extension could add the possibility to experiment with many novel effects.
\\\par
Klingbeil~\cite{klingbeil2005software} describes the techniques used in a high quality graphical
spectral editing and resynthesis program called SPEAR\@. The FFT analysis is done taking into
account the fundamental frequency of the sound to analyze, so that the transform can be optimized
to accurately capture as many partials as possible. Spectral peaks are then selected and very quiet
peaks are discarded. A linear prediction method is used to follow the peaks over time. The data
structure for storage of spectral data involves division of time into a number of frames which keep
track of which partials are active during that time, so that the search for a specific partial can
be narrowed. User interface design choices are also covered in detail.

@@

As the Mellin transform can be implemented simply as an extension to the Fourier transform, it
should be fairly easy to add support for it in ScPy, and it will hopefully enable the use of
some different audio effects than can be achieved easily with the Fourier transform. However, the
paper mentions that a short-time Mellin transform has not been attempted, so some further work may
need to be done in that area first.

@@

Collins et al.~\cite{collins2003live} present an introduction to the field of live coding, where
code to generate sound is written live during a performance, often on a laptop computer. Although
the use graphical programming environments such as Reaktor, Pure Data, or MAX/MSP in live
performances is in some sense live coding, the paper focuses on the use of text based programming
environments. Many of the pros and cons of live coding are presented. Although live coding adds
some very interesting new possibilities to live music performance, it is not without it's
challenges. Case studies of two live coding environments are included: Slub and SuperCollider with
the just in time library.

ScPy can be used for live coding of audio effects, similarly to how the environments studied in
this paper can be used to live code synths and pattens. More details about how ScPy was designed to
facilitate this can be found in \autoref{sec:methodology}.
\\\par
Domain-Specific Languages (DSLs) are small programming languages designed for use in a specific
problem domain. They often lack general purpose programming constructs and have a syntax which is
very concise for describing programs in their domain. Van Deursen et al.~\cite{van2000domain}
present a survey of available DSL literature, along with the pros and cons of using a DSL, example
DSLs, design methodology and implementation approaches. The biggest advantage of DSLs is that they
enable expression of solutions in a way that is common for the problem domain. However, they can be
costly to develop and difficult to design. Implementation techniques include the classical approach
of interpretation or compilation, embedded languages / domain-specific libraries, preprocessing, or
compiler extension. Additionally, the references of the paper each include a short summary.

@@

Use of ScPy is through the classes \texttt{Py}, \texttt{PyOnce}, \texttt{PyFile}, and
\texttt{PyOnceFile}. \texttt{Py} is used for processing real-time data, and \texttt{PyOnce} is used
for doing initialization or other one time processing. Both have one required argument, a block of
Python code. The \texttt{PyFile} and \texttt{PyOnceFile} variants take a filename instead of inline
code. For all classes, a map of variable names may optionally be provided to bind SuperCollider
variables to Python variables. Additionally, there is an optional \texttt{DoneAction} argument
which specifies an action to occur in SuperCollider after the Python code has finished executing.

@@

Since UGens may only be used as part of a synth, \texttt{PyOnce} is provided as a wrapper which
hides this detail from the user and allows the execution of Python code outside of a synth. It
internally just starts a synth with the \texttt{doneAction} argument set such that the synth
terminates after running once.

@@

Throughout~\autoref{sec:lit}, articles are discussed which contain processing techniques that could
be used in ScPy. Although implementations of many of these techniques are likely to become a part
of ScPy in the future, there is no obstacle to the reader implementing them as Python libraries
usable by ScPy or even as contributions to the project.

\section{Conclusion}

While the number of mentioned future extensions to ScPy is exhausting to consider, the basic
implementation that currently exists is more than enough to yield countless exotic and novel audio
effects. In the limited experiments that have been done using the software, it was found to be very
easy to produce many advanced sounds, thanks to the excellent documentation and implementation of
the Python language and NumPy library. Although it was certainly possible to make these effects
before, ScPy makes it possible to experiment with them at an unprecedented rate and listen to their
output in real-time.

Although initial tests have been promising, they have so far have had a focus on spectral
processing, leaving time domain audio and control rate data processing largely untested. Further
experiments in these areas could reveal previously unknown new use cases for the project.

However, the code may contain bugs due to the infancy of the project and the nonstandard ways we are
interfacing with SuperCollider. Users are encouraged to submit bug reports on the project's GitHub
page\footnote{\url{https://github.com/nwoeanhinnogaehr/ScPy}}.

\printbibliography{}
